---
title: "Data manipulation with dplyr"
author: "Susie Irons"
date: "21/01/2021"
output:
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 1 - Transforming data with dplyr

### Section 1 - The counties dataset

#### 2015 USA Census
- County = subregion of a state
- Glimpse allows you to see a few observations in all columns

```{r}
library('dplyr')
library('tigris')
counties
glimpse(counties)
```

#### Select
- Extracts only particular variables from a dataset

```{r}
counties %>%
  select(name, total.pop)
```

#### Creating a new table
- Assignment can be used to create a new table

```{r}
counties_select <- counties %>%
  select(name, total.pop)
counties_select
```

### Section 2 - The filter and arrange verbs

#### Arrange
- Orders selected variable from smallest to largest
- Useful finding the most interesting observations in the dataset

```{r}
counties_select %>%
  arrange(total.pop)
```
- Arrange in descending order

```{r}
counties_select %>%
  arrange(desc(total.pop))
```

#### Filter
- Extract only particular observations from a dataset, based on a condition
- Note the code below won't run as the dataset doesn't contain a 'state' column
- Can filter on logicals (<, >, <=, >=)
- Can combine multiple conditions in a filter by separating with a comma

```{r}
counties_select %>%
  arrange(desc(total.pop)) %>%
  filter(state == 'New York')
```

### Section 3 - Mutate
- The mutate verb can be used to add new variables or change existing ones

```{r}
counties_select %>%
  mutate(unemployed_population = population * unemployment / 100)
```

## Chapter 2 - Aggregating data

### Section 1 - The count verb

#### Count
- Counts total data
```{r}
counties %>%
  count()
```

- Total counties in each state
```{r}
counties %>%
  count(state)
```

#### Count and sort
- Often combine with sort to find most common observations
```{r}
counties %>%
  count(state, sort=TRUE)
```
- Sorts in desc order

#### Add weight
```{r}
counties %>%
  count(state, wt = popluation, sort=TRUE)
```
- Adds up the population of each county

### Section 2 - The group by, summarise and ungroup verbs

#### Group by and summarise
- Count is a case of a more general set of verbs
- Summarise takes many observations and reduces to a single figure

```{r}
counties %>%
  summarize(total_population = sum(population))
```

#### Aggregate and summarise
- Can define multiple variables
- Can aggregate in different ways

- Summary functions:
  - sum()
  - mean()
  - median()
  - min()
  - max()
  - n() - group size

#### Aggregate within groups

```{r}
counties %>%
  group_by(state) %>%
  summarize(total_pop = sum(population),
            average_unemployment = sum(unemployment)) %>%
  arrange(desc(average_unemployment))
```

- Useful to add an arrange to focus on most notable examples
- You can group by multiple columns at the same time

```{r}
counties %>%
  group_by(state, metro) %>%
  summarize(total_pop = sum(population))
```

- If you don't want to keep state as a group you can add ungroup()
```{r}
counties %>%
  group_by(state, metro) %>%
  summarize(total_pop = sum(population)) %>%
  ungroup()
```

#### Section 3 - The top_n verb

- Useful for keeping the most extreme observations from each group
- Operates on a grouped table
- Take number of obs and column want to weigh by

```{r}
counties_selected <- counties %>%
  select(state, county, population, unemployment, income)

counties_selected %>%
  group_by(state) %>%
  top_n(1, population)
```
- Finds the county with the highest population in each state
- n can be any number
- Often used when creating graphs - when interested in pulling extreme examples

## Chapter 3 - Selecting

### Selecting
#### Select a range
```{r}
counties %>%
  select(state, county, drive:work_at_home) %>%
  arrange(drive)
```

#### Contains
- Select helpers - functions that specify criteria for choosing columns
- Take strings - in quotes
```{r}
counties %>%
  select(state, county, contains("work"))
```

#### Starts with
- Selects only cols that start with a particular prefix
```{r}
counties %>%
  select(state, county, starts_with("income"))
```

#### Other helpers
- contains()
- starts_with()
- ends_with()
- last_col()

For more - `? select helpers`

#### Removing a variable
```{r}
counties %>%
  select(-census_id)
```

### Section 2 - The rename verb

```{r}
counties_selected <- counties %>%
  select(state, county, population, unemployment)

counties_selected %>%
  rename(unemployment_rate = unemployment)
```

#### Combine verbs
- Can be done as a part of select

```{r}
counties_selected %>%
  select(state, county, population, unemployment_rate = unemployment)
```

### Section 3 - The transmute verb

#### Transmute
- Combination - select & mutate
- Returns a subset of columns that are transformed and changed

```{r}
counties %>%
  transmute(state, county, fraction_men = men / population)
```

#### Summary

Properties          Keeps only specified variables    Keeps other variables
----------          -------------------------------   ----------------------
Can't change values  select                            rename
Can change values   transmute                          mutate

## Chapter 4 - Case study - the babynames dataset

```{r}
library('babynames')
babynames = babynames %>%
  select(year, name, n)
babynames
```

#### Frequency of a name

```{r}
babynames %>%
  filter(name == "Susannah")
```

#### Susannah plot
```{r}
library(ggplot2)

babynames_filtered <- babynames %>%
  filter(name == 'Susannah') %>%
  rename(number = n)
babynames_filtered

ggplot(babynames_filtered,
       aes(x = year, y = number, color = name)) +
      geom_line()
```
#### Filter for multiple names

```{r}
babynames_multiple <- babynames %>%
  filter(name %in% c("Susannah", "Patrick")) %>%
  rename(number = n)

ggplot(babynames_multiple,
       aes(x = year, y = number, color = name)) +
      geom_line()
```

#### When was each name most common?

```{r}
babynames %>%
  group_by(name) %>%
  top_n(1, n)
```
### Section 2 - Grouped mutates
```{r}
babynames %>%
  group_by(year) %>%
  mutate(year_total = sum(n))
```

- Group by - only want to add up within each year
- Mutate - creates a new column called year_total
- Presence of group_by can make other groups or filters slower to run
- Good practice to use ungroup() - as we've done what we're trying to do
- Can then calculate the fraction column

```{r}
babynames %>%
  group_by(year) %>%
  mutate(year_total = sum(number)) %>%
  ungroup() %>%
  mutate(fraction = number / year_total)
```

- Fractions can make graphs more interpretable

### Section 3 - Window functions

#### Window function
- Takes a vector and returns a vector of the same length
- Using the `lag()` function - moves each item to the right
- This allows for comparison between consecutive steps and calculating the changes

```{r}
v <- c(1, 3, 6, 14)
v
lag(v)
v - lag(v)  # Difference between two observations
```

#### Changes in popularity of a name

```{r}
babynames_fraction <- babynames %>%
  group_by(year) %>%
  mutate(year_total = sum(n)) %>%
  ungroup() %>%
  mutate(fraction = n / year_total)

babynames_fraction
```
#### Susannah

```{r}
babynames_fraction %>%
  filter(name == 'Susannah') %>%
  arrange(year)
```
#### Susannah over time
```{r}
babynames_fraction %>%
  filter(name == "Susannah") %>%
  arrange(year) %>%
  mutate(difference = fraction - lag(fraction))
```
#### Biggest jump in popularity
```{r}
babynames_fraction %>%
  filter(name == "Susannah") %>%
  arrange(year) %>%
  mutate(difference = fraction - lag(fraction)) %>%
  arrange(desc(difference))
```
#### Changes within every name
```{r}
babynames_fraction %>%
  arrange(name, year) %>%
  mutate(difference = fraction - lag(fraction)) %>%
  group_by(name) %>%
  arrange(desc(difference))
```

