---
title: "Joining data with dplyr"
output: html_notebook
---

## Chapter 1 - Joining tables

### Section 1 - The inner_join verb
- The LEGO dataset
- Import dplyr
```{r}
library('dplyr')
```


#### Import the sets table
```{r}
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
sets <- read.csv("sets.csv")
sets
```

#### Import the themes table
```{r}
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
themes <- read.csv("themes.csv")
themes
```

#### Linking two tables
- Inner join
```{r}
sets %>%
  inner_join(themes, by = c("theme_id" = "id"))
```

- Joins sets to themes
- by - tells inner join to join theme_id in first table to id in second table
- Cannot have two variables with the same name 
- Without specifying x and y will be used

#### Customizing your join 
```{r}
sets %>%
  inner_join(themes, by = c("theme_id" = "id"), suffix = c("_set", "_theme"))
```

- Appends _set or _theme to the columns - makes it much more readable

#### Most common themes

```{r}
sets %>%
  inner_join(themes, by = c("theme_id" = "id"), 
             suffix = c("_set", "_theme")) %>%
  count(name_theme, sort = TRUE)
```
#### Import parts
```{r}
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
parts <- read.csv("parts.csv")
parts

```
#### Import part_categories
```{r}
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
part_categories <- read.csv("part_categories.csv")
part_categories
```

### Section 2 - Joining with a one-to-many relationship
#### Import inventories
```{r}
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
inventories <- read.csv("inventories.csv")
inventories

```

#### Joining sets and inventories
- When variables have the same name, can just write the name

```{r}
sets %>%
  inner_join(inventories, by = "set_num") %>%
  filter(version != 1)
```
- Number of observations increases - each set can have multiple versions
- Each gets own inventory item
- Can filter by versions

- Inner join - keeps observation only if it has an exact match between tables
- Important to pay attention to the number of rows before and after joining

#### Inventory parts
- Combines inventory and color
```{r}
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
inventory_parts <- read.csv("inventory_parts.csv")
inventory_parts
```

- Combination of color and part describes a *piece*
- For an `inner_join`, either direction will yield a table that contains the same info - but columns will appear in a different order

### Section 3 - Joining three or more tables
- Same approach applies
```{r}
sets %>%
  inner_join(inventories, by = "set_num") %>%
  inner_join(themes, by = c("theme_id" = "id"),
             suffix = c("_set", "_theme"))
```
- In a series of joins - each will have different by arguments and they may have different suffixes

## Chapter 2 - Left and Right Joins

### Section 1 - The `left_join` verb
```{r}
inventory_parts_joined <- inventories %>%
  inner_join(inventory_parts, by = c("id" = "inventory_id")) %>%
  select(-id, -version) %>%
  arrange(desc(quantity))

inventory_parts_joined
```

- Select to remove id and version variables
- Sorting in desc order of quantity

#### Filter for batmobile and batwing
```{r}
batmobile <- inventory_parts_joined %>%
  filter(set_num == "7784-1") %>%
           select(-set_num, -is_spare)

batwing <- inventory_parts_joined %>%
  filter(set_num == "70916-1") %>%
  select(-set_num, -is_spare)

batmobile
batwing
```
- Each observation is a lego piece a combination of part and color

#### Joining with multiple columns
- Inner join - can find out what pieces are in common
- Joining needs to be by piece - a combination of part and color

```{r}
batmobile %>%
  inner_join(batwing, by = c("part_num", "color_id"),
             suffix = c("_batmobile", "_batwing"))
```
#### The left join
- Keeps parts that are in the batmobile, but not in the batwing
- Inner join - only keeps obs that appear in both tables
- If want to keep all the obs in one table - use left join

```{r}
batmobile %>%
  left_join(batwing, by = c("part_num", "color_id"), 
            suffix = c("_batmobile", "_batwing"))
```
- The batwing column will now have NAs in it, where they don't appear in the batwing set

- Left join - keep all the observations in the left table, whether or not it occurs in the right table

### Section 2 - The right-join verb
- Opposite of a left-join

#### Count and sort
```{r}
sets %>%
  count(theme_id, sort=TRUE) %>%
  inner_join(themes, by = c("theme_id" = "id"))
```
- Missing themes that never occurred in the dataset
- Right join would keep themes that never occur
```{r}
sets %>%
  count(theme_id, sort=TRUE) %>%
  inner_join(themes, by = c("theme_id" = "id"))
```

#### Replace NAs
- From tidyr package
```{r}
library(tidyr)

sets %>%
  count(theme_id, sort=TRUE) %>%
  right_join(themes, by = c("theme_id" = "id")) %>%
  replace_na(list(n = 0))
```
- Replaces any NA in n column

### Section 3 - Joining tables to themselves
- Hierarchical tables - parent_id often used to represent hierarchical relationships
- To explore the relationship between layers of a hierarchical table, can join to itself

#### Child-parent table
```{r}
themes %>%
  inner_join(themes, by = c("parent_id" = "id"),
             suffix = c("_child", "_parent"))
```
#### Lord of the Rings themes: parent
```{r}
themes %>%
  inner_join(themes, by = c("parent_id" = "id"),
             suffix = c("_child", "_parent")) %>%
  filter(name_child == "The Lord of the Rings")
```
#### Lord of the Rings themes: children
```{r}
themes %>%
  inner_join(themes, by = c("parent_id" = "id"),
             suffix = c("_child", "_parent")) %>%
  filter(name_parent == "The Lord of the Rings")
```

- Cominbing these helps us to understand the shape of our data

## Chapter 3 - Full, semi and anti-joins

### Section 1 - The `full_join` verb
- Keeps all observations in both tables, whether or not they match to each other

#### Joining batmobile and batwing together
- Left join - keep all batmobile
- Right joint - keep all batwing
- Full join - keeps all observations in both
```{r}
batmobile %>%
  full_join(batwing, by = c("part_num", "color_id"), 
            suffix = c("_batmobile", "_batwing"))
```
- n = 440 - more than either table individually

#### Replace NA: multiple variables
```{r}
batmobile %>%
  full_join(batwing, by = c("part_num", "color_id"), 
            suffix = c("_batmobile", "_batwing")) %>%
  replace_na(list(quantity_batmobile = 0,
                  quantity_batwing = 0))
```
list

### Section 2 - The semi- and anti- joins
#### Mutating verbs
- `inner_join`
- `left_join`
- `right_join`
- `full_join`

- When joining - you end up with new cols you didn't have before

#### Filtering joins
- Keeps or removes observations from the first table
- Doesn't add new variables
- `semi_join()`
- `anti_join()`

- Semi join
  - What observations in X are also in Y
- Anti join
  - What observations in X are not in Y
  
#### Semi join
- Useful for filtering a table without modifying
- Don't need a suffix
```{r}
batmobile %>%
  semi_join(batwing, by = c("color_id", "part_num"))
```

#### The anti join
- Also don't need a suffix
```{r}
batmobile %>%
  anti_join(batwing, by = c("color_id", "part_num"))
```
### Section 3 - Visualizing set differences

#### Aggregating sets into colors
- Batmobile
```{r}
batmobile_colors <- batmobile %>%
  group_by(color_id) %>%
  summarize(total = sum(quantity))

batmobile_colors
```
- Batwing
```{r}
batwing_colors <- batwing %>%
  group_by(color_id) %>%
  summarize(total = sum(quantity))

batwing_colors
```

#### Comparing color schemes of sets
- Combine both tables into single table and replace NAs
```{r}
all_colors <- batmobile_colors %>%
  full_join(batwing_colors, by = "color_id", suffix = c("_batmobile", "_batwing")) %>%
  replace_na(list(total_batmobile = 0, total_batwing = 0))

all_colors
```
#### Adding the color names
- Using inner join
```{r}
# Import colors
setwd("C:\\Users\\susie\\Documents\\R Projects\\Datacamp\\lego_data")
colors <- read.csv("colors.csv")
colors

# Join to all_colors
all_colors %>%
  inner_join(colors, by = c("color_id" = "id"))
```
#### Normalize each of the colors - turn into fractions of total
```{r}
all_colors %>%
  inner_join(colors, by = c("color_id" = "id")) %>%
  mutate(total_batmobile = total_batmobile / sum(total_batmobile),
         total_batwing = total_batwing / sum(total_batwing))
```
#### The difference between fractions
- Relative difference between colors in each set
```{r}
colors_joined <- all_colors %>%
  inner_join(colors, by = c("color_id" = "id")) %>%
  mutate(total_batmobile = total_batmobile / sum(total_batmobile),
         total_batwing = total_batwing / sum(total_batwing),
         difference = total_batmobile - total_batwing)

colors_joined
```
#### Visualiseing the data

```{r}
# Not working - # missing from from of colors
library(ggplot2)
library(forcats)

color_palette <- setNames(colors_joined$rgb, colors_joined$name)

colors_joined %>%
  mutate(name = fct_reorder(name, difference)) %>%
  ggplot(aes(name, difference, fill = name)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = color_palette, guide = FALSE)
```

## Chapter 4 - Case study - joins on stack overflow data
### Section 1 - Stackoverflow questions
- Intermediate table - links IDs, no actual tables
- Common in web databases
### Section 2 - The bind_rows verb
- In some situations - may want to stack tabes on top of one another
```{r}
questions %>%
  bind_rows(answers)
```
- When combining tables - you can mutate in an extra col that distinguishes each of them before combining
```{r}
questions_type <- questions %>%
  mutate(type = "question")

answers_type <- answers %>%
  mutate(type = "answer")

posts <- bind_rows(questions_type, answers_type)
posts
```

#### Aggregating
- Could then use table for aggregating

#### Creating date variable
```{r}
library(lubridate)

posts %>%
  mutate(year = year(creation_date))
```

